When does vapor pressure deficit drive or reduce evapotranspiration?

### 2019-09-05 doc/paper/agu-james-submission.tex
submitting resubmission now

The ET response to changes in VPD alters water partitioning between
the soil and atmosphere. If ecosystem plant response reduces ET with
atmospheric drying then soil moisture will be better conserved. This
represents a sensible evolutionary strategy to cope with aridity: save
water for periods when atmospheric demand for water is relatively low,
and atmospheric carbon can be accessed with a relatively smaller cost
in water loss. If instead stomata were fully passive \cite<similar to
soil pores, e.g. >[]{Or_2013}, increased atmospheric aridity would
strongly reduce soil moisture \cite{Berg_2017}. This could further
increase aridity as low soil moisture levels increases the Bowen
ratio, leading to increased temperature and atmospheric drying
\cite{Bouchet_1963, Morton_1965, Brutsaert_1999, Ozdogan_2006,
  Salvucci_2013, Gentine_2016, Berg_2016, Zhou_2019}. Therefore,
passive regulation and a lack of soil moisture conservation does not
seem to be a sensible strategy for plants from an evolutionary
standpoint. This simplified logic explains generally why plants
evolved to respond to VPD, but also excludes many details and special
cases \cite<e.g. plant to plant interaction, thermal regulation with
transpiration, and highly specialized photosynthesis strategies like
Crassulacean acid metabolism photosynthesis>[]{Brooker_2006, Lin_2017,
  Cushman_2001}.

@@

  where g$_1$ is presented units of $\sqrt{\text{mol fraction}}$ for
  convenience (which can be converted to $\sqrt{Pa}$ using the ideal
  gas law and ambient pressure), $\Gamma^*$ is the CO$_2$ compensation
  point for photosynthesis (without dark respiration), and $\lambda$
  is the marginal water cost of leaf carbon
  ($\frac{\partial \; \text{transpiration}}{\partial \; CO_2
    assimilation}$) \cite{Farquhar_1980, Katul_2009}. So, $g_{1}$ is a
  leaf-scale term reflecting the trade-off of water for carbon
  uptake. The higher $g_{1}$, the more open the stomata and the more
  they release water in exchange for carbon.

@@

However, given that g$_1$ and uWUE can vary, and they both have a
large impact on the ET response (Figure \ref{full}, \ref{sign}), it is
useful to examine how these quantities vary in previous literature,
framed by our results on $\frac{\partial \; ET}{\partial \; VPD}$. In
\citeA{Zhou_2015} most plant types' uWUE is similar to our median
value, under which the sign of the ET response depends on
VPD. However, crops, shurbs, and grass are exceptions; they exhibit an
uWUE different than the median value in Table
\ref{param_varying}. Crops, which we expect to prioritize carbon
uptake over water conservation, have a higher uWUE value closer to our
maximum value. This higher uWUE results in a higher likelihood for an
increasing ET response to VPD, which matches intuition given that we
expect crops to keep stomata open for access to carbon, at the cost of
increased water loss during high VPD. Shrubs, and to a lesser extent
grass, have an uWUE closer to the minimum uWUE. For these plant types,
we would then expect a decrease of ET in response to VPD. It is
important to note that, while \citeA{Zhou_2015} did not examine the
role of soil moisture for within plant-type variability of uWUE, we
might expect some variation, especially in extreme cases when soil
water becomes a limiting factor (see supplementary material).

The ecosystem-scale results of \citeA{Medlyn_2017} did not display the
robust relationship between plant type and and plant parameter (g$_1$)
that was demonstrated by \citeA{Zhou_2015} for uWUE. Given that there
is a robust relationship between g$_1$ and plant type in the
leaf-scale results of \citeA{Lin_2015}, and the ecosystem-scale
results of \citeA{Zhou_2015} rely on the same underlying theory as
\citeA{Medlyn_2017}, the ambiguous results in \citeA{Medlyn_2017}
could be due to noise in ecosystem scale observations, the
consequences of imposing a model structure in some of their g$_1$
estimations, or real ecosystem-scale within-plant type variability
(e.g. multiple species impacting the relationship within the
eddy-covariance footprint). These differences between leaf-scale
behavior and ecosystem scale behavior highlight the importance of
understanding how and why slope parameters show increased variability
at the ecosystem scale, as argued by \citeA{Medlyn_2017}. However, if
we assume that there is some real analogy between leaf scale results
in \citeA{Lin_2015} and ecosystem scale behavior, and that some of the
ambiguity in \citeA{Medlyn_2017} is due to model and/or observational
error, then we can use the relationships between g$_1$ and plant
characteristics defined by \citeA{Lin_2015} to frame the ET response
to VPD. This is subject to the strong caveat that we still do not
fully understand g$_1$ behavior at the ecosystem scale. Extrapolating
the results of \citeA{Lin_2015} (e.g. Fig. 2) to the ecosystem scale
would reveal the following relationships between g$_1$ and plant
types:

@@

\begin{figure}
  \centering
  \centerline{\includegraphics[width=0.75\textwidth]{./concave.pdf}}
  \caption{ Solutions corresponding to inflection points between
    concave up and concave down ET-VPD curves (Eq. (\ref{curves})) for
    three specific scenarios. Solutions are defined in terms of a
    non-dimensional VPD ($VPD^m/g_*$), but to aide physical
    interpretation the horizontal axis is additionally provided in
    terms of dimensionalized VPD assuming $m=1/2$ and
    $g_*=110\; Pa^{1/2}$ (average of all PFT $g_1$). The vertical axis
    has a different interpretation depending on the solution
    curve. For the blue line ($m$ varying), it corresponds to $m$, for
    the orange line ($n$ varying) it corresponds to $n$, and for the
    green line it corresponds to the value of both $n$ and $m$
    ($n=m$). Regions of the parameter space that correspond to concave
    up and concave down results are labeled: for each curve, the
    parameter space ``below'' the line corresponds to a concave up
    relationship, which the parameter space ``above'' the line
    corresponds to a concave down relationship.}
  \label{concave}
\end{figure}

>>>>

The ET response to changes in VPD alters water partitioning between
the soil and atmosphere. If ecosystem plant response reduces ET with
atmospheric drying then soil moisture will be better conserved. This
represents a sensible evolutionary strategy to cope with aridity: save
water for periods when atmospheric demand for water is relatively low,
and atmospheric carbon can be accessed with a relatively smaller cost
in water loss. If instead stomata were fully passive \cite<similar to
soil pores, e.g. >[]{Or_2013}, increased atmospheric aridity would
strongly reduce soil moisture \cite{Berg_2017}. This could further
increase aridity as low soil moisture levels increases the Bowen
ratio, leading to increased temperature and atmospheric drying
\cite{Bouchet_1963, Morton_1965, Brutsaert_1999, Ozdogan_2006,
  Salvucci_2013, Gentine_2016, Berg_2016, Zhou_2019}. Therefore,
passive regulation and a lack of soil moisture conservation does not
seem to be a sensible strategy for plants from an evolutionary
standpoint. This simplified logic explains generally why plants
evolved to respond to VPD, but also excludes many details and special
cases \cite<e.g. plant to plant interaction, thermal regulation with
transpiration, and highly specialized photosynthesis strategies like
Crassulacean acid metabolism photosynthesis; see>[]{Brooker_2006, Lin_2017,
  Cushman_2001}.

@@

  where g$_1$ is presented in units of $\sqrt{\text{mol fraction}}$ for
  convenience (which can be converted to $\sqrt{Pa}$ using the ideal
  gas law and ambient pressure), $\Gamma^*$ is the CO$_2$ compensation
  point for photosynthesis (without dark respiration), and $\lambda$
  is the marginal water cost of leaf carbon
  ($\frac{\partial \; \text{transpiration}}{\partial \; CO_2
    assimilation}$) \cite{Farquhar_1980, Katul_2009}. So, $g_{1}$ is a
  leaf-scale term reflecting the trade-off of water for carbon
  uptake. The higher $g_{1}$, the more open the stomata and the more
  they release water in exchange for carbon.

@@

However, given that g$_1$ and uWUE can vary, and they both have a
large impact on the ET response (Figure \ref{full}, \ref{sign}), it is
useful to examine how these quantities vary in previous literature,
framed by our results on $\frac{\partial \; ET}{\partial \; VPD}$. In
\citeA{Zhou_2015} most plant types' uWUE is similar to our median
value, under which the sign of the ET response depends on
VPD. However, crops, shrubs, and grass are exceptions; they exhibit an
uWUE different than the median value in Table
\ref{param_varying}. Crops, which we expect to prioritize carbon
uptake over water conservation, have a higher uWUE value closer to our
maximum value. This higher uWUE results in a higher likelihood for an
increasing ET response to VPD, which matches intuition given that we
expect crops to keep stomata open for access to carbon, at the cost of
increased water loss during high VPD. Shrubs, and to a lesser extent
grass, have an uWUE closer to the minimum uWUE. For these plant types,
we would then expect a decrease of ET in response to VPD. It is
important to note that, while \citeA{Zhou_2015} did not examine the
role of soil moisture for within plant-type variability of uWUE, we
might expect some variation, especially in extreme cases when soil
water becomes a limiting factor (see supplementary material).

The ecosystem-scale results of \citeA{Medlyn_2017} did not display the
robust relationship between plant type and and plant parameter (g$_1$)
that was demonstrated by \citeA{Zhou_2015} for uWUE. Given that there
is a robust relationship between g$_1$ and plant type in the
leaf-scale results of \citeA{Lin_2015}, and the ecosystem-scale
results of \citeA{Zhou_2015} rely on the same underlying theory as in
\citeA{Medlyn_2017}, the ambiguous results in \citeA{Medlyn_2017}
could be due to noise in ecosystem scale observations, the
consequences of imposing a model structure in some of their g$_1$
estimations, or real ecosystem-scale within-plant type variability
(e.g. multiple species impacting the relationship within the
eddy-covariance footprint). These differences between leaf-scale
behavior and ecosystem scale behavior highlight the importance of
understanding how and why slope parameters show increased variability
at the ecosystem scale, as argued by \citeA{Medlyn_2017}. However, if
we assume that there is some real analogy between leaf scale results
in \citeA{Lin_2015} and ecosystem scale behavior, and that some of the
ambiguity in \citeA{Medlyn_2017} is due to model and/or observational
error, then we can use the relationships between g$_1$ and plant
characteristics defined by \citeA{Lin_2015} to frame the ET response
to VPD. This is subject to the strong caveat that we still do not
fully understand g$_1$ behavior at the ecosystem scale. Extrapolating
the results of \citeA{Lin_2015} (e.g. Fig. 2) to the ecosystem scale
would reveal the following relationships between g$_1$ and plant
types:

@@

\begin{figure}
  \centering
  \centerline{\includegraphics[width=0.75\textwidth]{./concave.pdf}}
  \caption{ Solutions corresponding to inflection points between
    concave up and concave down ET-VPD curves (Eq. (\ref{curves})) for
    three specific scenarios. Solutions are defined in terms of a
    non-dimensional VPD ($VPD^m/g_*$), but to aide physical
    interpretation the horizontal axis is additionally provided in
    terms of dimensionalized VPD assuming $m=1/2$ and
    $g_*=110\; Pa^{1/2}$ (average of all PFT $g_1$). The vertical axis
    has a different interpretation depending on the solution
    curve. For the blue line ($m$ varying), it corresponds to $m$, for
    the orange line ($n$ varying) it corresponds to $n$, and for the
    green line it corresponds to the value of both $n$ and $m$
    ($n=m$). Regions of the parameter space that correspond to concave
    up and concave down results are labeled: for each curve, the
    parameter space ``below'' the curve corresponds to a concave up
    relationship, while the parameter space ``above'' the curve
    corresponds to a concave down relationship.}
  \label{concave}
\end{figure}
